---
title: "Converting Graph Classes"
author: "Brendan Knapp"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The two primary graph classes in R's network analysis ecosystem are `igraph` and `network`. While similar in many ways, the procedures provided do not always overlap. This is a challenge as the `igraph` and `network` classes are not immediately compatible, but analysts often require routines from both.

The common solution comes from the {`intergraph`} package, but the `intergraph::asIgraph()` and `intergraph::asNetwork()` methods have limitations.

There are two issues that {`snatools`} seeks to address with graph conversion via its `as_igraph()` and `as_network()` methods.

1. Enable safe, consistent conversion of all possible graph types, including bipartite graphs
2. Enable strict comparison of an initial graph and its state following conversion.
    + Manually inspecting every aspect of an object _does not replace built-in safety_.
    + Manual inspections should be possible, but be enabled by simple, built-in methods.


```{r, message=FALSE, warning=FALSE}
library(snatools)
library(igraph)
library(network)
```

